Longitudinal Principal Manifold Estimation

### 2023-12-14 paper/main.tex
Updates from Overleaf

Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability as measured by the Limits of Agreement was much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained. In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\%, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.

@@

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits. Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, but estimated parameterizations for baseline images and longitudinal image differences separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework. In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus in individuals with Alzheimer's disease (AD). The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings requires is problematic. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, i.e. $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows. 

@@

The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements. First is the expected distance from each point in the cloud to the fitted manifold. In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on 3-dimensional brain region of interest. While the second element of the equation imposes the smoothness penalty on the estimated manifold. The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting before. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$. In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. It is important to note, that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.

@@

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. In other words, we consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally, during several visits by a study participant. Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit. Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.

>>>>

Failures of automatic segmentation algorithms may have drastic effects when analyzing large datasets of images where manual quality control may not be feasible, such as data from the Alzheimer's Disease Neuroimaging Initiative (ADNI) study. In a large comparison of the accuracy of FIRST and FreeSurfer, \cite{mulderHippocampalVolumeChange2014} found that 6.9\% of segmentations by FIRST failed visual inspection for accuracy, as did 7.5\% of segmentations by FreeSurfer. After removing these failed segmentations, FIRST and FreeSurfer produced segmentations with variability similar to and slightly lower than manual segmentation, respectively. If failed segmentations were not removed from analysis, reflecting a more realistic situation when working with a large number of images, variability was much higher for FIRST and FreeSurfer than for manual segmentation. \cite{mulderHippocampalVolumeChange2014} also found slightly higher rates of segmentation failure among individuals with Alzheimer's disease for both automated segmentation approaches, suggesting that variability increases as images are taken from individuals with greater deviations from the healthy brains which compose the majority of the data on which the segmentation models were trained. In studying longitudinal trajectories of atrophy, segmentation problems may result in seeming random increases and reductions in brain region volumes, although it has been shown that brain volumes decrease as a result of normal aging (\cite{scahill2003longitudinal}). To anecdotally demonstrate the potential extent of the inaccuracies introduced by segmentation, using FIRST to segment the hippocampus in images of one healthy individual over the course of 6 visits during four years in the ADNI dataset, the volume of the left and right hippocampus decreased by 3.6\% and 5.4\% over the study duration, respectively, but each hippocampus was estimated to have increased in volume between subsequent study visits twice.

@@

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

Previous work on longitudinal dimension reduction methods has primarily focused on linear approaches, such as the longitudinal principal component analysis method proposed by \cite{kinsonLongitudinalPrincipalComponent2020}. \cite{wolzManifoldLearningBiomarker2010a} used Laplacian eigenmaps to incorporate longitudinal information when reducing neuroimaging data to lower dimensional space, but estimated parameterizations for baseline images and longitudinal image differences separately while we seek to parameterize them simultaneously. \cite{louisRiemannianGeometryLearning2019} used a deep learning-based approach to learn Riemannian manifolds in the context of disease progression modeling. However, this approach relies on modeling assumptions specific to the application, and the deep learning methodology used imposes costs in terms of computational requirements and interpretability. Our work includes two major contributions to the existing literature: 1) we propose a novel approach for longitudinal smoothing in the general-purpose manifold learning setting, 2) we show the potential of leveraging longitudinal manifold smoothing for reducing variability and increasing signal-to-noise in segmentation of brain structures over time.

The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework. In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus and thalamus in individuals with Alzheimer's disease. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

The framework for principal manifolds originated with the concept of principal curves (PS, \cite{hastiePrincipalCurves1989}), which are essentially curves that pass through the middle of a data cloud. However, as pointed out by \cite{duchamp1996extremal}, the PS framework in \cite{hastiePrincipalCurves1989} has limitations such as the model complexity selection. In addition, applying the methods in high-dimensional settings is problematic. Motivated by the penalization approaches presented by \cite{kegl2000learning} and \cite{smolaRegularizedPrincipalManifolds2001}, \cite{mengPrincipalManifoldEstimation2021} proposed the principal manifold estimation (PME) framework. In order to present the longitudinal PME approach, we start by introducing notations used throughout this paper. We use $d$ and $D$ to denote the dimensions of the low-dimensional manifold and high-dimensional space, respectively. We assume that a data cloud is observed in the $D$ dimensional space. In the context of segmentation, we consider a collection of vertices on the surface of the brain region of interest as the $D-$dimensional data cloud, while the $d-$dimensional smooth surface of the region is the underlying manifold of interest. For any positive integer $q$ and point $\xi=(\xi_1,\ldots,\xi_q)\in\mathbb{R}^q$, we use the following norm $\Vert \xi\Vert_{\mathbb{R}^q}=\sqrt{\sum_{l=1}^q \xi_l^2}$. Since our interest in this work centers at continuous mappings, we denote the collection of continuous vector-valued functions from $\mathbb{R}^d$ to $\mathbb{R}^D$ as $\mathcal{C}(\mathbb{R}^{d} \to \mathbb{R}^{D})$. The $d$-dimensional manifold corresponding to a function $f \in C(\mathbb{R}^{d} \to \mathbb{R}^{D})$ is formally defined as $M_f^d=\{f(r):\,r\in\mathbb{R}^d\}$. Most manifold estimation methods are based on defining a projection index, i.e. $\pi_f(x)$, defined as the point in the $d$-dimensional space, where $f(\pi_f(x))$ is the projection of $x$ on the manifold $M_f^d$. The definitions of $\pi_f(x)$ given by \cite{hastiePrincipalCurves1989} and \cite{mengPrincipalManifoldEstimation2021} for $2$- and higher-dimensional spaces, respectively, guarantee the uniqueness of the projection. The principal manifolds are then defined as follows. 

@@

The optimization function $\mathcal{K}_{\lambda, \mathbb{P}}$ in equation (\ref{eq:5}) consists of two elements. First is the expected distance from each point in the cloud to the fitted manifold. In our application, this would correspond to minimizing the distance between the fitted surface and the coordinates of the vertices on the 3-dimensional brain region of interest. The second element of the equation imposes the smoothness penalty on the estimated manifold. The choice of $\|f''(r)\|_{\mathbb{R}^{D}}^2$ provides a greater penalization of roughness than using the first derivative often implemented in this setting before. To allow for manifolds with intrinsic dimension $d \geq 1$ and map $f(t) = \left(f_1(r), f_2(r), \dots, f_D(r)\right)^{T}$, this term can be generalized using the total squared curvature of the manifold and defined as $\|\nabla^{\otimes 2}f\|_{L^2(\mathbb{R}^{d})}^2 = \sum_{l=1}^{D} \int_{\mathbb{R}^{d}}\sum_{i, j = 1}^{d}\left|\frac{\partial^2f_l}{\partial r_i \partial r_j}(r)\right|^2dr$. In this setting, we can see that $\pi_{f_{\lambda}^{*}}(\boldsymbol{X})$ maps the $D$-vector $\boldsymbol{X}$ to a $d$-dimensional parameterization, while $f_{\lambda}^{*}$ embeds the $d$-dimensional parameterization in the original $D$-dimensional space. It is important to note that the  magnitude of the penalty on roughness in this framework depends on the given value of $\lambda$.

@@

The framework described above is an excellent approach for fitting a 2-dimensional principal surface that passes through the ``middle'' of a 3-dimensional cloud of vertices. We first heuristically explain the goal of our proposed model and its importance in segmentation of brain regions for longitudinal imaging studies. We consider the problem of reducing the dimension of the coordinates $(x_1, x_2, x_3)$, i.e. $D=3$ to $(y_1,y_2)$ with $d=2$, which are the projections of the original coordinates on the surface of the 3-dimensional shape of interest. Now consider that the data are observed longitudinally during several visits by a study participant. Hence, the data of interest consists of the 4-dimensional cloud $(x_1, x_2, x_3,t_j)$, where $t_j$ denotes the time of the $j$th study visit. Figure \ref{fig:sample_data} displays a simplified example of this setting using simulated data where $D = 2$. The goal is to fit a surface through the data that depends on time and is not affected by the random noise added at each time point and in space.
